---
title: "Getting started with prettysheets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with prettysheets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`prettysheets` extends [`googlesheets4`](https://googlesheets4.tidyverse.org/)
with a formatting layer for Google Sheets: cell formatting, borders, merged
cells, frozen rows/columns, tab colors, conditional formatting, and reusable
table themes. Every function reuses whatever auth `googlesheets4` already
has active, and function names/argument order mirror `googlesheets4`'s own
conventions (`gs4_*` spreadsheet-level, `sheet_*` worksheet-level, `range_*`
range-level), so it behaves like a natural extension rather than a separate
tool.

Every example below talks to a real Google Sheet, so it isn't run when this
vignette is built. Copy the code into a live R session with a Google account
already authenticated (or about to be, via `googlesheets4::gs4_auth()`) to
try it.

```{r setup}
library(googlesheets4)
library(prettysheets)

ss <- gs4_create("prettysheets-demo", sheets = list(cars = mtcars))
```

## Basic cell formatting

`range_format()` is `prettysheets`'s equivalent of
`googlesheets4::range_write()` for styling rather than data: same `ss`,
`sheet`, `range` argument order, and it accepts the same range inputs (A1
strings or `cellranger` helpers).

```{r}
range_format(ss, sheet = "cars", range = "A1:K1", bold = TRUE, background_color = "#D9E1F2")
range_format(ss, sheet = "cars", range = "B2:B33", number_format = "0.0", horizontal_alignment = "RIGHT")
```

Repeated calls layer on top of each other rather than overwriting -- only
the arguments you actually supply are touched. Pass `clear_first = TRUE` to
reset a range to blank before applying new formatting instead.

Colors can be a hex string, a named R color, or a name from the built-in
Google Sheets color picker grid (`gs4_palette`, looked up via
`gs4_palette_color()`):

```{r}
range_format(ss, sheet = "cars", range = "H3:I4", background_color = "dark yellow 1")
gs4_color("light cornflower blue 1")
```

Other structural helpers follow the same pattern:

```{r}
range_format_border(ss, sheet = "cars", range = "A1:K33", sides = "outside", color = "gray40")
range_merge(ss, sheet = "cars", range = "A1:B1")
sheet_freeze(ss, sheet = "cars", n_rows = 1)
sheet_format_header(ss, sheet = "cars") # bold header + freeze, for data written with range_write()
sheet_format_tabcolor(ss, sheet = "cars", color = "forestgreen")
sheet_clear_format(ss, sheet = "cars") # reset everything, including banding/conditional formats
```

## Referencing columns by name

`gs4_cols()` targets one or more columns by their header name instead of a
column letter -- handy when you don't want to count columns by hand, and
robust to a sheet's columns being reordered later.

```{r}
range_format(ss, sheet = "cars", range = gs4_cols("mpg", "hp"), number_format = "0.0")
```

If the named columns aren't contiguous, this expands into one `GridRange`
per contiguous run under the hood; the header row itself is excluded by
default (pass `include_header = TRUE` to include it).

## Conditional formatting

Boolean rules are built with `cf_cell_value()` (or `cf_text_contains()`,
`cf_text_eq()`, `cf_custom_formula()`, `cf_blank()`, `cf_not_blank()`) and a
`cf_format()`, then applied with `range_add_conditional_format()`:

```{r}
range_add_conditional_format(
  ss, sheet = "cars", range = gs4_cols("mpg"),
  rule = cf_cell_value(">", 25),
  format = cf_format(bold = TRUE, background_color = "#F4CCCC")
)
```

Gradient (color-scale) rules use `cf_gradient()` and
`range_add_gradient_format()`:

```{r}
range_add_gradient_format(
  ss, sheet = "cars", range = gs4_cols("hp"),
  gradient = cf_gradient(min_color = "white", max_color = "forestgreen")
)
```

Conditional formatting only supports five cell properties in the Sheets
API -- bold, italic, strikethrough, and foreground/background color --
which is why `cf_format()` doesn't expose number formats or alignment.

## Write and format in one call

`range_write_format()` combines `googlesheets4::range_write()` with these
formatting functions, and -- unlike `range_write()` itself -- creates the
target sheet if it doesn't already exist, sized to fit the data:

```{r}
combined_data <- data.frame(x = c(5, 15, 25, 35), y = letters[1:4])

range_write_format(
  ss, sheet = "combined", data = combined_data,
  header_background_color = "#D9E1F2",
  conditional_formats = list(
    list(range = "A2:A5", rule = cf_cell_value(">", 20),
         format = cf_format(background_color = "#F4CCCC"))
  )
)
```

## Reusable table themes

For a consistent table look applied in one call, `gs_theme()` bundles
header style, zebra-striping (banding), per-column overrides, a border, and
conditional formats into one reusable, inspectable object:

```{r}
my_theme <- gs_theme(
  header = list(background_color = "#1F4E79", font_color = "white"),
  banding = list(band1 = "white", band2 = "#DCE6F1"),
  columns = list(mpg = list(number_format = "0.0")),
  border = list(color = "gray40")
)
my_theme
```

`write_pretty_sheet()` is the theme-aware counterpart to
`range_write_format()`: it writes a data frame and applies a theme in one
call, creating the sheet first if needed.

```{r}
write_pretty_sheet(ss, data = mtcars, sheet = "themed", theme = my_theme)
```

`theme` can also be the name of a built-in preset (`"clean"`,
`"professional"`, `"fun"`, or `"stylish"`), or omitted entirely to fall back
to a plain default. `modify_theme()` tweaks an existing theme without
rebuilding it from scratch:

```{r}
tweaked <- modify_theme(gs_theme_preset("professional"), banding = list(band2 = "lightyellow"))
```

The four presets also apply immediately to an already-written sheet, for
when you don't need a reusable theme object:

```{r}
sheet_write(mtcars, ss, sheet = "mtcars2") |> gs_theme_professional()
```

## Reading a sheet's theme back

`read_pretty_sheet()` is the theme-aware counterpart to
`googlesheets4::range_read()`: it reads a sheet's data and reconstructs a
`gs_theme()` describing how the table is formatted (header style, banding,
per-column formatting, and any conditional-format rules), attached to the
returned tibble as `attr(x, "prettysheets_theme")`. That makes it easy to
copy a sheet's look elsewhere:

```{r}
styled <- read_pretty_sheet(ss, sheet = "themed")
write_pretty_sheet(ss, data = styled, sheet = "themed_copy") # same theme, no `theme` argument needed
```

## Batching multiple calls into one request

Every `prettysheets` function normally sends its own `batchUpdate` call as
soon as it runs. Wrap a block of calls in `prettysheets_batch()` to collect
them and send everything together in a single request instead, which
matters if you're making many formatting calls in a row and want to stay
comfortably under the Sheets API's rate limits:

```{r}
prettysheets_batch({
  range_format(ss, sheet = "cars", range = "A1:K1", bold = TRUE)
  range_format_border(ss, sheet = "cars", range = "A1:K1", color = "gray40")
})
```
